### 2015-06-17 chapters/ipa_meets_unicode.tex
update to homoglyphs in IPA section

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard.

\noindent Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
the following.
\begin{itemize}

  \item The uses of the apostrophe has led to long discussions on the Unicode
     Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
     at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
     \textsc{right single quotation mark} at \uni{2019}.\footnote{Note that many word 
     processors (like Microsoft Word) by default will replace straight quotes by 
     curly quotes, depending on the whitespace around it. Such a replacement is 
     a real character replacement, meaning that another Unicode code point is 
     inserted.} However, the glottal stop/glottalization/ejective marker is yet 
     another completely different character <{\large ʼ}>, the 
     \textsc{modifier letter apostrophe} at \uni{02BC}, which unfortunately looks 
     mostly extremely similar to \uni{2019}. 
  \item Another problem is the <ˁ> \textsc{modifier letter reversed
     glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
     letter small reversed glottal stop} at \uni{02E4}. Both 
     appear in various resources representing phonetic data online. This is
     thus a clear example for which the Unicode Standard does not solve the
     linguistic standardization problem.
  \item even linguists are unlikely to distinguish between the <ə>
     \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
     \textsc{latin small letter turned e} at \uni{01DD}.
  \item  The alveolar click <ǃ> at \uni{01C3} is of course often simply
     typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the Unicode
     Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter retroflex
     click}, but in IPA that glyph is used for an alveolar or postalveolar
     click (not retroflex). This naming is probably best seen as an 
     error in the Unicode Standard.}

@@

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter a}
at \uni{0061}, basically treating them as homoglyphs, although they are
different phonetic symbols.

But even among linguists this distinction leads to problems. For example, as
pointed out by \citet{Mielke2009}, there is a problem stemming from the fact
that about 75\% of languages are reported to have a five-vowel system
\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
precise audio recording and measurements of formants, so this may lead one to
ask if the many <a> characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions could
be in part due to the ease of typing the letter on an English keyboard (or for
older descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The diacritic for voiceless
  \item The tie bar to indicate a close bond between sounds
\end{itemize}

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
for high tone.} The IPA also proposes a less widespread used, but more
consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
five different levels, and sequences of these letters can be used to indicate
contours. Well-designed fonts will even merge a sequence of tone letters into a
contour. For example, compare the font Linux Libertine, which does not merge
tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
For strict IPA encoding we propose to standardize on tone letters.

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. This issue was recently addressed by The International Phonetic
Association has taken the stance that both the keyboard \textsc{latin small
letter g} and the \textsc{latin small letter script g} are valid input
characters for the voiced velar plosive. Unfortunately, this decision further
introduces ambiguity for linguists trying to adhere to a strict Unicode Standard
IPA encoding. For strict IPA encoding we propose to standardize on the more 
idiosyncratic \textsc{latin small letter script g} at \uni{0261}.

>>>>

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard. Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
the following.
\begin{itemize}

  \item The uses of the apostrophe has led to long discussions on the Unicode
     Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
     at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
     \textsc{right single quotation mark} at \uni{2019}.\footnote{Note that many word 
     processors (like Microsoft Word) by default will replace straight quotes by 
     curly quotes, depending on the whitespace around it.} However, the glottal stop/glottalization/ejective marker is yet 
     another completely different character <{\large ʼ}>, the 
     \textsc{modifier letter apostrophe} at \uni{02BC}, which unfortunately looks 
     mostly extremely similar to \uni{2019}. 
  \item Another problem is the <ˁ> \textsc{modifier letter reversed
     glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
     letter small reversed glottal stop} at \uni{02E4}. Both 
     appear in various resources representing phonetic data online. This is
     thus a clear example for which the Unicode Standard does not solve the
     linguistic standardization problem.
  \item even linguists are unlikely to distinguish between the <ə>
     \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
     \textsc{latin small letter turned e} at \uni{01DD}.
  \item  The alveolar click <ǃ> at \uni{01C3} is of course often simply
     typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the Unicode
     Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter retroflex
     click}, but in IPA that glyph is used for an alveolar or postalveolar
     click (not retroflex). This naming is probably best seen as an 
     error in the Unicode Standard.}

@@

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter a}
at \uni{0061}, basically treating them as homoglyphs, although they are
different phonetic symbols. But even among linguists this distinction leads to problems. For example, as
pointed out by \citet{Mielke2009}, there is a problem stemming from the fact
that about 75\% of languages are reported to have a five-vowel system
\citep{Maddieson1984}. Historically, linguistic descriptions tend not to include
precise audio recording and measurements of formants, so this may lead one to
ask if the many <a> characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions could
be in part due to the ease of typing the letter on an English keyboard (or for
older descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The placement of diacritics
  \item The tie bar to indicate a close bond between sounds
\end{itemize}

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
complexe, there are at least two different Unicode homoglyphs for the low and
high level tones, namely <\ ̀~> \textsc{combining grave tone mark} at
\unif{0340} vs.\@ <\ ̀~> \textsc{combining grave accent} at \unif{0300} for low
tone, and <\ ́~> \textsc{combining acute tone mark} at \unif{0341} vs.\@ <\ ́~>
\textsc{combining acute accent} at \unif{0301} for high tone.} The IPA also
proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are used much less, but
are more consistent for contours. Tone letters in the IPA have five different levels, and
sequences of these letters can be used to indicate contours. Well-designed fonts
will even merge a sequence of tone letters into a contour. For example, compare
the font Linux Libertine, which does not merge tone letters
<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
IPA encoding we propose to standardize on tone letters.

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. The International Phonetic
Association has taken the stance that both the keyboard \textsc{latin small
letter g} and the \textsc{latin small letter script g} are valid input
characters for the voiced velar plosive. Unfortunately, this decision further
introduces ambiguity for linguists trying to adhere to a strict Unicode Standard
IPA encoding. For strict IPA encoding we propose to standardize on the more 
idiosyncratic \textsc{latin small letter script g} at \uni{0261}.
